---
title       : West Indian population in CT
author      : Scott Gaul
job         : Community Indicators Project
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
mode        : selfcontained # {standalone, draft}
markdown    : kramdown
---

## What is the rate of new arrivals from West Indies?

Metro Hartford has the highest rate of immigration from the West Indies in the United States ([naturalizations and new permanent residents](http://www.dhs.gov/data-statistics)). About 1 in 6 new arrivals for the region are from the West Indies. (This excludes immigration from Cuba, the Dominican Republic and Haiti.) 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 7, fig.width = 14}
#Read in files, remove na strings
#Immigration data from here: http://www.dhs.gov/yearbook-immigration-statistics-2011-naturalizations 
library(reshape)
library(plyr)
library(scales)
library(ggplot2)
immstats <- rbind(
  melt(read.csv('immsuptable2d-2012.csv',na.strings = c("-","D")),
       id.vars = c("Year","Status","Country")),
  melt(read.csv('immsuptable2d-2011.csv', na.strings = c("-","D")),
       id.vars = c("Year","Status","Country")),
  melt(read.csv('natzsuptable2d-2012.csv', na.strings = c("-","D")),
       id.vars = c("Year","Status","Country")),
  melt(read.csv('natzsuptable2d-2011.csv', na.strings = c("-","D")),
       id.vars = c("Year","Status","Country")))

imm.west.indies <- 
  merge(ddply(immstats,.(Status, variable), summarise,
            total = sum(value, na.rm = T)),
      ddply(subset(immstats, Country %in% c("Jamaica")),
            .(Status, variable), summarise,
            west.indies = sum(value, na.rm = T)),
      by = c("Status","variable"))

imm.west.indies <- 
  merge(ddply(immstats,.(Status, variable), summarise,
            total = sum(value, na.rm = T)),
      ddply(subset(immstats, Country %in% c("Bahamas","Barbados","Jamaica","Dominica","Grenada","St. Vincent and the Grenadines","Trinidad and Tobago","Guyana","Antigua-Barbuda","Anguilla","Aruba","British Virgin Islands","Cayman Islands","Netherlands Antilles","Saint Kitts-Nevis","Saint Lucia","Turks and Caicos Islands")),
            .(Status, variable), summarise,
            west.indies = sum(value, na.rm = T)),
      by = c("Status","variable"))

imm.west.indies$percent.all = imm.west.indies$west.indies / imm.west.indies$total

imm.west.indies$variable <- reorder(imm.west.indies$variable, imm.west.indies$percent)

ggplot(data = subset(imm.west.indies, 
                     variable %in% head(arrange(imm.west.indies, desc(percent.all)),n = 10)$variable)) + 
  geom_bar(aes(x = variable, y = percent.all, fill = Status), stat ="identity", position = "dodge") + 
  labs(x = NULL, y = "Percent from West Indies (2011 - 12)") +
  coord_flip() + 
  scale_y_continuous(labels = percent) +
  theme_bw(base_size = 14) 
```

--- .class #id

Looking only at arrivals from Jamaica, the Hartford metro region still has the highest percent nationally. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 7, fig.width = 14}
imm.west.indies <- 
  merge(ddply(immstats,.(Status, variable), summarise,
            total = sum(value, na.rm = T)),
      ddply(subset(immstats, Country %in% c("Jamaica")),
            .(Status, variable), summarise,
            west.indies = sum(value, na.rm = T)),
      by = c("Status","variable"))

imm.west.indies$percent.all = imm.west.indies$west.indies / imm.west.indies$total

imm.west.indies$variable <- reorder(imm.west.indies$variable, imm.west.indies$percent)

ggplot(data = subset(imm.west.indies, 
                     variable %in% head(arrange(imm.west.indies, desc(percent.all)),n = 10)$variable)) + 
  geom_bar(aes(x = variable, y = percent.all, fill = Status), stat ="identity", position = "dodge") + 
  labs(x = NULL, y = "Percent from Jamaica (2011 - 12)") +
  coord_flip() + 
  scale_y_continuous(labels = percent) +
  theme_bw(base_size = 14) 
```

--- .class #id

Within the Caribbean, arrivals from Jamaica are about 80% of all arrivals for 2011-12. Guyana and St. Lucia are the next countries with the most immigrants to metro Hartford. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 7, fig.width = 12}
hartford.imm <- subset(immstats,
                       variable == "Hartford.West.Hartford.East.Hartford..CT" & 
                         Country %in% c("Bahamas","Barbados","Jamaica","Dominica","Grenada","St. Vincent and the Grenadines","Trinidad and Tobago","Guyana","Antigua-Barbuda","Anguilla","Aruba","British Virgin Islands","Cayman Islands","Netherlands Antilles","Saint Kitts-Nevis","Saint Lucia","Turks and Caicos Islands") & !(is.na(value)))

hartford.imm$Country <- reorder(hartford.imm$Country, hartford.imm$value)

ggplot(data = subset(hartford.imm, value > 0 & Year == 2012)) + 
    geom_bar(aes(x = Country, y = value, fill = Status), stat ="identity", position = "dodge") + 
  labs(x = NULL, y = "Number of immigrants") +
  scale_y_continuous(labels = comma) + 
  facet_wrap(~ Year) + 
  coord_flip() +
  theme_bw(base_size = 14) 

#Get percent Jamaican of immigrants
#ddply(subset(hartford.imm, Country == "Jamaica"), 
#      .(), summarise, value = sum(value,na.rm = T))[2] / 
#ddply(hartford.imm, .(), summarise, value = sum(value,na.rm = T))[2]
```

--- .class #id

## Where does the West Indian population live now? 

People of Caribbean origin live in many towns in the state, although mainly the urban centers. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=6, fig.width = 12, fig.align='center'}
#Load the UConn town-level shapefiles
#Load the UConn shapefile
library(maps)
library(maptools)
CTTowns <- readShapeSpatial(fn="../commute/townshp/wgs84/townct_37800_0000_2010_s100_census_1_shp_wgs84")

#Fortify and order the CTTowns to allow ggmap to use that
CTTowns <- fortify(CTTowns, region = "NAME10")
CTTowns <- CTTowns[order(CTTowns$order),]

#Merge with data
choropleth=merge(CTTowns, B05006.town, by.x = "id", by.y="geo")
choropleth=choropleth[order(choropleth$order), ]
choropleth$newcarib=cut(choropleth$newcarib, 
                       breaks=c(0,100,500,1000,5000,20000),
                       include.lowest=T, dig.lab = 5)
#Make the map
ggplot(data = choropleth, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = newcarib)) + 
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(breaks = NULL) +
  labs(x = NULL, y = NULL) + 
  coord_equal() +
  geom_polygon(data = CTTowns, colour = "grey", alpha = 0.1, size = 0.1, fill = NA) +
  scale_fill_brewer(palette = "Purples", name = "Born in\nAnglo Caribbean") +
  theme_minimal() 
```

--- .class #id

The highest percent from West Indies and Jamaica in the state are in Bloomfield, followed by Hartford and Windsor. East Hartford also has a large West Indian population.

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=6, fig.width = 12, fig.align='center'}
#Create towns for the state
ct.towns = geo.make(state = "CT", county = c("Hartford","Tolland","New Haven","New London","Windham","Fairfield","Litchfield","Middlesex"), county.subdivision = "*", check = F)

#Percent of foreign-born population from Caribbean
B05006 = acs.fetch(geography = ct.towns, table.number = "B05006", col.names = "pretty", year = 2012)

#Total population
B01003 = acs.fetch(geography = ct.towns, table.number = "B01003", col.names = "pretty", year = 2012)

B05006.town = data.frame(geo=geography(B05006)[[1]],
                          caribbean = as.numeric(estimate(B05006[,124])),
                          dr = as.numeric(estimate(B05006[,129])),
                          haiti = as.numeric(estimate(B05006[,131])),
                          jamaica = as.numeric(estimate(B05006[,132])),
                         guyana = as.numeric(estimate(B05006[,154])))

B01003.town = data.frame(geo=geography(B01003)[[1]],
                          population = as.numeric(estimate(B01003[,1])))

B05006.town = merge(B05006.town, B01003.town, by = "geo")

B05006.town$geo= gsub(" town, (Hartford|New Haven|Middlesex|Litchfield|Fairfield|Windham|New London|Tolland) County, Connecticut", "", B05006.town$geo)

B05006.town$newcarib = B05006.town$caribbean - B05006.town$dr + B05006.town$guyana
B05006.town$pct.carib = B05006.town$newcarib / B05006.town$population
B05006.town$pct.jamaica = B05006.town$jamaica / B05006.town$population

B05006.town$geo <- reorder(B05006.town$geo, B05006.town$pct.carib)

ggplot(data = subset(B05006.town, pct.carib > 0.02)) + 
  geom_point(aes(y = geo, x = pct.carib), colour = "blue", size = 3) + 
  geom_point(aes(y = geo, x = pct.jamaica), colour = "green", size = 3) + 
  labs(x = "Percent Caribbean / Jamaican", y = NULL) + 
  scale_x_continuous(labels = percent) + 
  theme_bw(base_size = 14)

#Totals: subset(B05006.town, pct.carib > 0.03)[c(1,9,10)]
```

--- .class #id

The Jamaican population are mostly clustered in the smaller set of towns around Hartford. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=6, fig.width = 12, fig.align='center'}
#And then for Jamaica
choropleth=merge(CTTowns, B05006.town, by.x = "id", by.y="geo")
choropleth=choropleth[order(choropleth$order), ]
choropleth$pct.jamaica=cut(choropleth$pct.jamaica, 
                       breaks=c(0,.02,.05,.1,.25),
                       include.lowest=T, dig.lab = 5)
#Make the map
ggplot(data = choropleth, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pct.jamaica)) + 
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(breaks = NULL) +
  labs(x = NULL, y = NULL) + 
  coord_equal() +
  geom_polygon(data = CTTowns, colour = "grey", alpha = 0.1, size = 0.1, fill = NA) +
  scale_fill_brewer(palette = "Purples", name = "Born in\nJamaica") +
  theme_minimal() 
```

--- .class #id
## Are there students that speak Patois at home? 

Some Connecticut students are recorded as speaking Patois (or 'Patwa') as a second language at home. For the state, Patois was recorded as a home language for about 200 students in 2013, or about the same as the occurence of Karen and Khmer. About [30,000 Connecticut residents](http://factfinder2.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_12_1YR_B05006&prodType=table) were born in Jamaica, half living in metro Hartford.

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=6, fig.width = 14, fig.align='center'}
ell <- read.csv('NON ENGLISH LANGUAGES BY DISTRICT 2011-13.csv',
                na.strings = c("<=5"))

#Clean up the languages that sound like mistakes
ell$Language <- str_trim(ell$Language)
ell$Language <- gsub(pattern = "Patwa", replacement = "Patois", ell$Language)
ell$Language <- gsub(pattern = "Creole-Jaotoam", replacement = "Creole-Haitian", ell$Language)
ell$Language <- gsub(pattern = "Kiswhali", replacement = "Swahili", ell$Language)
ell$Language <- gsub(pattern = "Rawandan", replacement = "Kinyarwanda", ell$Language)
ell$Language <- gsub(pattern = "Tagalog", replacement = "Pilipino (Filipino)", ell$Language)

library(stringr)
ell$District <- str_trim(ell$District, side = "both")

ell13 <- subset(ddply(ell,.(Language, School.Year),summarise,
                           Students = sum(Students, na.rm = T),
                           ELLs = sum(ELLs, na.rm = T)),
                     Students > 100 & Language != "Spanish" & School.Year == "2013-14")
ell13$Language <- reorder(ell13$Language, ell13$Students)

ggplot(data = ell13,
       aes(y = Students, x = Language)) + 
  geom_bar(width = 0.5, stat = "identity") + 
  labs(x = NULL, y = "Students speaking language at home (excl. Spanish)") + 
  coord_flip() +
  theme_bw(base_size = 14)
```

--- .class #id

Some students from multilingual households are recorded as English-language learners. In this case, about 30 students were recorded as Patois-speakers in need of English language assistance in the state. (A few districts have fewer than 5 students and don't report exact results.)

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=6, fig.width = 14, fig.align='center'}
ell13$Language <- reorder(ell13$Language, ell13$ELLs)

ggplot(data = ell13,
       aes(y = ELLs, x = Language)) + 
  geom_bar(width = 0.5, stat = "identity") + 
  coord_flip() +
  labs(x = NULL, y = "Home language for English-language learners (excl. Spanish)") + 
  theme_bw(base_size = 14)
```

--- .class #id
As a percent of total district enrollment, Patois speakers only comprise a significant share at Jumoke Academy, with between 15 - 20% of students from Patois-speaking households. This is similar to the base population rates in Hartford, Bloomfield and Windsor. No other district in the region reports more than 2% students coming from a Patois-speaking household. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=6, fig.width = 14, fig.align='center'}
ell.enroll <- subset(merge(ell,
                           read.csv('enrollment.csv', na.strings = c("*")),
                           by.x = c("School.Year","District"),
                           by.y = c("School.Year","district.name")),
                     !(is.na(Students)))

ell.enroll$pct.students = ell.enroll$Students / ell.enroll$Total
ell.enroll$pct.ell = ell.enroll$ELLs / ell.enroll$Total

ggplot(data = ddply(subset(ell.enroll, 
                           Language %in% c("Patois","Patwa") & pct.students >0),
                    .(District,School.Year,Language), 
                    summarise,
                    pct.students = sum(pct.students, na.rm = T),
                    pct.ell = sum(pct.ell, na.rm = T))) + 
  geom_bar(aes(x = School.Year, y = pct.students), width =.5, stat="identity") + 
  scale_y_continuous(labels = percent) + 
    labs(x = "School Year", y = "Patois-speaking students as % of  enrollment") + 
  facet_wrap(~ District, nrow = 2) + 
    theme_bw(base_size = 14)
```

--- .class #id

In addition, for English-language learners, no school district has more than 1.25% students registered as Patois-speakers in need of English language assistance and only four districts (Achievement First Hartford, Hartford, Windsor and Bloomfield) register more than 5 students total. 

```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height=6, fig.width = 14, fig.align='center'}
ggplot(data = ddply(subset(ell.enroll, 
                           Language %in% c("Patois","Patwa") & pct.students >0),
                    .(District,School.Year,Language), 
                    summarise,
                    pct.students = sum(pct.students, na.rm = T),
                    pct.ell = sum(pct.ell, na.rm = T))) + 
  geom_bar(aes(x = School.Year, y = pct.ell), width =.5, stat="identity") + 
  scale_y_continuous(labels = percent) + 
  labs(x = "School Year", y = "ELL students as % of enrollment") + 
  facet_wrap(~ District, nrow = 2) + 
    theme_bw(base_size = 13)
```



